---
title: "Assignment-3\n \\vspace{1in}"
output: pdf_document
date: "2023-03-31\n \\vspace{1in}"
author: "Imanbayeva Sofya, Mazzi Lapo, Piras Mattia, Srivastava Dev\n \\vspace{1in}"
subtitle: "Group 22\n \\vspace{1in}"
---

```{r setup, include=FALSE, echo=TRUE}
knitr::opts_chunk$set(echo = FALSE, results = 'asis')
list.of.packages <- c("fpp2", "forecast","ggplot2","depmixS4","tidyverse","dplyr","rlang","tseries","zoo")
if(!require(fpp2)){
    install.packages("fpp2")
    library(fpp2)
}
if(!require(forecast)){
    install.packages("forecast")
    library(forecast)
}
if(!require(ggplot2)){
    install.packages("ggplot2")
    library(ggplot2)
}

if(!require(depmixS4)){
    install.packages("depmixS4")
    library(depmixS4)
}

if(!require(tidyverse)){
    install.packages("tidyverse")
    library(tidyverse)
}

if(!require(dplyr)){
    install.packages("dplyr")
    library(dplyr)
}

if(!require(rlang)){
    install.packages("rlang")
    library(rlang)
}
  if(!require(tseries)){
    install.packages("tseries")
    library(tseries)
  }
  if(!require(zoo)){
    install.packages("zoo")
    library(zoo)
}
```
\newpage


```{r loading dataset}
airdataset=read.csv("https://raw.githubusercontent.com/devpra1999/Time-Series-20236/main/ts_epa_2020_west_sept_fill.csv",header = TRUE)
attach(airdataset)
st55datset= subset(airdataset,station_id==55)
st55datset$Time <- as.POSIXct(st55datset$datetime, format = "%Y-%m-%dT%H:%M:%SZ")
## had some troubles with Time, the previous one with T%hh resulted in NA values, had to get rid of hours, but maybe there's a better solution 
attach(st55datset)
```

```{r time series pm25}
pm25ts <- zoo(st55datset$pm25, st55datset$datetime)
daily_avg_pm25 <- aggregate(pm25ts, as.Date(index(pm25ts)), mean)
daily_avg_ma_pm25 <- rollmean(daily_avg_pm25, k = 24, fill = NA)
plot(daily_avg_ma_pm25,main = "Graph 1: Average PM25 levels at station 55, 2020", xlab = "Time", ylab= "Average PM25 level per day")

```
Moving average dramatically increases in the beginning of August from around 17 to its peak in the beginning of September at around 60 in its PM25 levels. The values slightly decrease in September.

```{r ts temp}
tempts <- zoo(st55datset$temp, st55datset$datetime)
daily_avg_t <- aggregate(tempts, as.Date(index(tempts)), mean)
daily_avg_ma_t <- rollmean(daily_avg_t, k = 24, fill = NA)
plot(daily_avg_ma_t, main = "Graph 2: Average Daily Temperature (C°) at station 55, 2020", xlab = "Time", ylab= "Average Daily Temperature (C°)")
```
The temperature has been increasing from June's values of 24 degrees Celsius to its peak in August at around 28.

```{r ts wind}
windts <- zoo(st55datset$wind, st55datset$datetime)
daily_avg_w <- aggregate(windts, as.Date(index(windts)), mean)
daily_avg_ma_w <- rollmean(daily_avg_w, k = 24, fill = NA)
plot(daily_avg_ma_w, main = "Graph 3: Average Wind Speed (knots/second) at station 55, 2020", xlab = "Time", ylab= "Average Wind Speed (knots/second)")
```
Average wind level has been stable at around 14m/s in the summer and started to decrease in the second part of August to around 12 in mid-September.

```{r compare}
plot(scale(daily_avg_ma_pm25), col = "darkgreen", main = "Graph 4: Scaled development of PM25, temperature and wind, 2020", cex.main=0.9,xlab = "Time", ylab= "Average Temperature level per day")
lines(scale(daily_avg_ma_w), col = "blue")
lines(scale(daily_avg_ma_t), col = "red")
legend("bottomleft", legend = c("PM25","Wind", "Temp"), col = c("darkgreen","blue","red"), lty = c(1,1,1), cex = 0.5)
```
High average temperature in August and comparatively strong winds seem to have a correlation with fires, which have increased the values of PM25 particles in the air. There is around 10 day lag between the temperature increases and PM25 value increase in August.

```{r PM25 levels}
xmin = min(st55datset$Time)
xmax = max(st55datset$Time)
ymax = max(st55datset$pm25)

p <- ggplot(data=st55datset, aes(x=Time, y=pm25)) +
  ggtitle("PM2.5 levels at Station #55") +
  geom_rect(aes(xmin=xmin, xmax=xmax, ymin=25, ymax=ymax), fill=adjustcolor("pink",alpha.f=0.9), alpha=.2) +
  geom_line() +
  annotate(geom="text", x=as.POSIXct("2020-06-18"), y=200, label="Dangerous PM2.5 level", color="darkred") +
  geom_hline(yintercept=25, color="darkred") + 
  scale_y_continuous(expand=c(0,0)) +
#  scale_x_date(limit=c(as.POSIXct("2020-06-01 00:00:00"),as.POSIXct("2020-09-30 23:00:00")))+
  labs(x=NULL, y=NULL)
p
```
Majority of measurements from June to mid-August are smaller than the prescribed limit with the exception of a peak of an outlying 307.81 in July. However, since then until October, the dynamic has changed with only a few days where the values staying within the limit constraints below 25. The peaks are high, probably resulted from fires, high temperatures and strong wind.

```{r wind levels}
### Time should be fixed and the average(?) of wind taken into account

 windy <- ggplot(data=st55datset, aes(x=Time, y=wind)) +
   ggtitle("Wind strength") +
   geom_line() + 
   scale_y_continuous(expand=c(0,0)) +
#   scale_x_date(limit=c(as.Date("2020-06-01"),as.Date("2020-09-30")))+
   labs(x=NULL, y=NULL)
 windy
```
First, we  set up our model. The initial probabilities and the transition matrix are just made by default values. Since, there are 3 states, they will have 1/3 probability each.

```{r model}
y=as.numeric(st55datset$pm25)
model = depmix(y~ 1, data=data.frame(y), nstates=3)
```

```{r, comment=""}

  cat('\n') 
  model


```


```{r MLE}
fmodel <- fit(model)

```
The results of the estimation of both the initial probabilities and the transition matrix are indicated below.
```{r,echo=TRUE}
summary(fmodel)
```
First thing looking at the mean for the parameters of the emissions distribution we can identify the three states that we wanted to study. State 2 is the one relative to low pollution, state 3 is relative to high pollution levels and state 1 is the one that we can associate to a medium pollution levels.
Looking at the transition matrix we note that there are steps that are never possible, such as the one from state 2 to 3.

```{r}
estStates <- posterior(fmodel)
```
```{r, out.width='40%'}
plot(as.POSIXct(st55datset$datetime), estStates[,1],ylab= "PREDICTED STATE",xlab="TIME",  cex=.3, type="s")

```
