---
title: "Assignment-5\n \\vspace{1in}"
output: pdf_document
date: "2023-04-26\n \\vspace{1in}"
author: "Imanbayeva Sofya, Mazzi Lapo, Piras Mattia, Srivastava Dev\n \\vspace{1in}"
subtitle: "Group 22\n \\vspace{1in}"
---
\newpage
```{r setup, include=FALSE, echo=TRUE}
knitr::opts_chunk$set(echo = FALSE, out.height = "75%", out.width = "75%")
list.of.packages <- c("dlm", "fpp2", "forecast","ggplot2","depmixS4","tidyverse","dplyr","rlang","tseries","zoo")
if(!require(dlm)){
    install.packages("dlm")
    library(dlm)
}
if(!require(fpp2)){
    install.packages("fpp2")
    library(fpp2)
}
if(!require(forecast)){
    install.packages("forecast")
    library(forecast)
}
if(!require(ggplot2)){
    install.packages("ggplot2")
    library(ggplot2)
}
if(!require(depmixS4)){
    install.packages("depmixS4")
    library(depmixS4)
}
if(!require(markovchain)){
    install.packages("markovchain")
    library(markovchain)
}
if(!require(tidyverse)){
    install.packages("tidyverse")
    library(tidyverse)
}
if(!require(dplyr)){
    install.packages("dplyr")
    library(dplyr)
}
if(!require(rlang)){
    install.packages("rlang")
    library(rlang)
}
  if(!require(tseries)){
    install.packages("tseries")
    library(tseries)
  }
  if(!require(zoo)){
    install.packages("zoo")
    library(zoo)
}
```

# Question 1 - Filtered Estimates

The following plot shows the annual flow in the river Nile, the data we are going to model using a dynamic linear model
```{r Nile graph}
par(mar = c(2, 4, 1, 1) + 0.1, cex = 0.6)
plot(Nile, main = "Graph 1: River Nile annual flow in 10^8 cubic meters (1871-1970)", xlab = "Time", ylab= "Water flow in 10^8 cubic meters")
```

Let us consider the following random walk plus noise model to be applied to this Nile data:
\begin{eqnarray*} 
Y_t &= \theta_t + v_t \quad    & v_t \overset{i.i.d.}\sim N(0, V)\\
\theta_t &= \theta_{t-1} + w_t \quad   & v_t \overset{i.i.d.}\sim N(0, W)
\end{eqnarray*}

We will set V = 15100 and W = 1470 and the initial distribution $\theta_0 \sim N(1000,1000)$ for our model

```{r model}
mod <- dlm(m0=1000, C0=1000, FF=1, V=15100, GG=1, W=1470)
NileFilt <- dlmFilter(Nile, mod)
```

Plotting the filtered estimates we get

```{r plot filtered estimates, out.height="200px"}
par(mar = c(2, 4, 1, 1) + 0.1, cex = 0.8)
plot(dropFirst(NileFilt$m),ylab = "Filtered Estimates", main = "Graph 2: Filtered Estimates using V=15100, W=1470")
```

We can compute the variance as well. It is plotted as follows
```{r variance compute,out.height="200px"}
par(mar = c(2, 4, 1, 1) + 0.1, cex = 0.8)
varC <- unlist(dlmSvd2var(NileFilt$U.C, NileFilt$D.C))
sdC <- sqrt(varC)
ts.plot(sdC[-1], ylab="Standard Deviation", main = "Graph 3: Standard Deviation of the filtered estimates")
```

As we can see from the graph above, the standard deviations of the filtering estimates decrease as time passes by, and converges to a value above 60. This might reflect the low guess of the standard deviation C0 (sqrt(1000) = 31) made at the beginning . As new observations arrive the standard deviation converges to the stable value. 


We finally plot the whole data along with the filtered estimates with 95% confidence interval

```{r final plot}
par(mar = c(2, 4, 1, 1) + 0.1, cex = 0.6)
plot(Nile, main = "Graph 4: Data and Filtered Estimates with 95% CI", xlab = "Time", ylab= "Water flow in 10^8 cubic meters", type = "o")
lines(dropFirst(NileFilt$m), col = "orange")
lines(dropFirst(NileFilt$m + 1.96*sdC), col = "blue", lty = "dotdash")
lines(dropFirst(NileFilt$m - 1.96*sdC), col = "blue", lty = "dotdash")
legend("bottomleft", legend = c("data","filtered","CI"),
col = c("black", "orange","blue"),
lty = c("solid", "solid", "dotdash"),
pch = c(16, NA, NA))
```



# Question 2 - One-step ahead forecasts


```{r compute sd for forecast}
varR <- unlist(dlmSvd2var(NileFilt$U.R,NileFilt$D.R))
sdR <- sqrt(varR)
```

Below is the plot for the one-step ahead forecast with 95% confidence intervals


```{r forecasting}
par(mar = c(2, 4, 1, 1) + 0.1, cex = 0.6)
a <- window(cbind(Nile, NileFilt$f),
start = 1871, end = 1970)
plot(a[, 1], type = 'o', col = "black",
xlab = "", ylab = "Level",main = "Graph 5: One Step ahead Forecasts with 95% CI")
lines(dropFirst(a[, 2]), lty = "solid", col='red')
lines(dropFirst(a[, 2] + 1.96*sdR), lty = "dotdash", col='blue')
lines(dropFirst(a[, 2] - 1.96*sdR), lty = "dotdash", col='blue')
leg <- c("data", "One-step-Ahead Forecast","CI")
legend("bottomleft", legend = leg,
col = c("black", "red","blue"),
lty = c("solid", "solid", "dotdash"),
pch = c(16, NA, NA))
```


# Question 3 - Signal to noise ratio


The signal-to-noise ratio are an important factor which set the weight of the most recent data point for the filtered estimates, with a larger value putting more weight to the recent data-point.

To test the effect of signal-to-noise ratio on our model we test the following models.
\linebreak

Model 1 (Signal-to-noise ratio = 0.097) -
\begin{eqnarray*} 
Y_t &= \theta_t + v_t \quad    & v_t \overset{i.i.d.}\sim N(0, 15100)\\
\theta_t &= \theta_{t-1} + w_t \quad   & v_t \overset{i.i.d.}\sim N(0, 1470)
\end{eqnarray*}

Model 2 (Signal-to-noise ratio = 10.27)-
\begin{eqnarray*} 
Y_t &= \theta_t + v_t \quad    & v_t \overset{i.i.d.}\sim N(0, 1470)\\
\theta_t &= \theta_{t-1} + w_t \quad   & v_t \overset{i.i.d.}\sim N(0, 15100)
\end{eqnarray*}


```{r Playing with signal to noise ratio}
mod1 <- dlmModPoly(order = 1, dV =15100 , dW =1470)
outFilt1 <- dlmFilter(Nile,mod1)

mod2 <- dlmModPoly(order = 1, dV =1470 , dW =15100)
outFilt2 <- dlmFilter(Nile,mod2)
```


Plotting the filtered estimates of both the models we find

```{r signal to noise graph}
par(mar = c(2, 4, 1, 1) + 0.1, cex = 0.9)
plot(Nile,type="o", xlab = "", ylab = "Level",col='black', main = "Graph 6: Signal-to-noise")
lines(dropFirst(outFilt1$m), lty = "longdash", col='red')
lines(dropFirst(outFilt2$m), lty = "dotdash", col='blue')
leg <- c("data", paste("filtered, W/V =",
format(c(W(mod1) / V(mod1),
W(mod2) / V(mod2)))))
legend("bottomright", legend = leg,
col=c("black", "red", "blue"),
lty = c("solid", "longdash", "dotdash"),
pch = c(1, NA, NA), bty = "n")

```

\linebreak

We can see from the graph that when we increase the signal to noise ratio (W/V) substantially, the filtered estimates follow the data more closely. This can be explained, because when we increase W/V we increase the weight on the most recent data point. In the plot, given we increase the ratio hundredfold the estimates almost match the most recent points.

